EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-10-26 BrainScience/response.letter.2.tex

\subsection*{\ovalbox{General Comments}}
This manuscript presents findings that contrast the performance of a selection of BCI classification methods in a single subject database. All in all, I find several aspects of this paper quite interesting, particularly its attempt to assess the impact of latency and component amplitude variation on classification performance. However, overall, I fear the range/scope of the paper is too limited to make a significant contribution to the literature.
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We would like to honestly thank the Reviewer for this absolutely excellent and detailed review.  We have modified the experiments and several parts of the manuscript in order to make it more suitable for publication.  The list of changes can be found at the end of this letter.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

1. Abstract, while automated analysis of EEG data certainly has significant potential with regards to the future of clinical EEG research, it is too soon to unqualifiedly state that these methods as they are "outshine the research done by the clinical EEG community." Some aspects of the introduction also veer towards hyperbole on occasion and may be better received if they were more conservatively stated.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We apologize for this mistake.  We definitely selected an incorrect wording because the message that we aimed to convey was exactly the opposite.  We have modified the abstract to emphasize the message and the idea that we failed to transmit in our original manuscript.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

2. Several notable papers in the domain of P300-BCI work should be referenced (e.g., Rakotomamonjy \& Guigue, 2008) as well as other prior efforts to evaluate P300-BCI classification methods (e.g., Manyakov, Chumerin, Combaz \& Hulle, 2011; Oweis, Hamdi, Ghazali \& Lwissy, 2013).

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
These important references were added to the manuscript.  Thank you for pointing them out.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

4. Page 12, Lines 383-394. It should be noted that the dataset from which the template ERP subject was drawn from was an ALS patient dataset. It is currently unclear why Subject Number 8 and trial number 2 was selected out of all the subjects and trials in the BNCI-Horizon database.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

This is a very important comment.  The reason we selected that particular trial and subject was because our manuscript stressed the importance of the waveform, the shape of the ERP.  The obtained P300 waveform of Subject Number 8, trial 2 was the most prototypical that we've found along the BNCI-Horizon dataset of ALS patients. The analysis was performed by visual observation.  We have now added also 4 more templates for the Active part of the pseudo-real dataset.

We have now included this reason on the manuscript which was mistakenly omitted on our first version.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

5. Integrating the template ERP signal into a Null-EEG stream and assuming everything is there except the P300 ERP component is probably an untenable assumption. As indicated in the manuscript itself, for instance, the null-EEG stream is a passive task. While the non-P300 trials in an active task contain additional activity associated with processes including target stimuli anticipation, active task focus, cognitive fatigue, etc. This is not true of the Null-EEG stream. Furthermore, combining and subsequently classifying EEG data from two different subjects is questionable at best given the physiological differences between the two. Differences in recording condition, equipment and task stimuli should also be considered. Collectively, these factors make it difficult to assess the value of the classification results observed on this study.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

The recording conditions of both datasets are exactly the same.  The EEG device, though not the same exact model, is from the same manufacturer and the sampling frequency is as close as possible. All the preprocessing stage was implemented according to~\citep{Riccio2013}.

Following Reviewer comments, the pseudo-real dataset is now divided in two.  First, we've replicated the experiment conducted by~\cite{Riccio2013} et al on 4 healthy subjects performing an active-modality P300 copy-spelling task while using the P300-Based BCI Speller. We also extended the original single null-EEG stream to 4 subjects while performing the passive-modality where the letter to copy-spell is not informed to the user and the subject is unaware of how the matrix works.

Regarding the active-modality, for each subject a the most prototypical P300 waveform template is selected and extracted.  This template is superimposed to each segment marked as hit on the EEG stream.  The baseline was obtained by picking randomly any no hit segment for the same subject.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

6. The separate manipulation of latency, as well as amplitude noise, is interesting and if extended may in and of itself prove to be an interesting manuscript in and of itself (e.g. contrasting the impact of a range of different latency variations and amplitude noise on BCI classification ability), true EEG data almost invariantly involves not only a mix of both noise sources but other sources such as cognitive fatigue, systematic task stimuli adaptation etc. Thus all three experiments represent assessments of isolated aspects of noise which would be interesting if expanded on in a more expansive simulation set (see point 4 and summary) but currently are too limited in scope to draw strong conclusions regarding the expected inter-component performance in real-world data.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

The purpose of these experiments were to assess the incidence of the most common source of noise in the generation of the P300, to the shape of the obtained waveform of the P300, "how the averaged signal looked-like".

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

It seems as though the entire study is built on the classification of a single active P300 BCI-Speller template from an ALS participant into a single null-EEG stream of a healthy subject performing a passive-viewing version of a BCI-Speller task. The generalizability of these results are thus difficult, if not impossible to assess. Given that assessments of classification methods for P300 BCI interfaces are fairly common (e.g. Manyakov et al., 2011; Oweis et al. 2013; Krusienski et al. 2010), it is unclear what the value of another contrast of classification methods is to the field. The assessment of the impact of sources of noise is interesting (see point 6), but would need to be expanded on considerably to be of practical value. Furthermore, while this manuscript frequently emphasizes the value of BCI to the clinical field. However, most clinicians are generally uninterested in classifying the presence/absence of a P300 response, arguably the most obvious cognitive ERP component extant. Instead, the focus in that area is instead on subtler dynamics in the EEG signal such as frequency slowing, triphasic waves, and generalized ictal activity. Indeed, the introduction of this manuscript mentions several EEG components of value which notably does not include the P300 in its list. If demonstrating clinical utility is of interest, it would seem more appropriate to target the classification of these subtler elements with a larger sample set and broader response ERP pool.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We would like to emphasize that the purpose of this work was to raise awareness about the automatic procedures that can analyze EEG signals by their waveforms, mimicking what traditionally EEG physicians performed on the clinical EEG.  As far as we are aware of, no work has addressed this topic in this form and we humbly believe it could be a contribution.  The Reviewer is right, but the objective of this manuscript was to establish the connection, and to focus on the goal of the BCI discipline to foster interaction between the clinical and BCI community.

We have modified the manuscript to emphasize these aspects.

\bibliographystyle{mdpi}
\bibliography{brainscience}

### 2018-10-26 BrainScience/brainscience.tex
Draft

The need of quantitative procedures to automate the decoding of EEG signals has been materialized in BCI where around $71.2\%$ is based on noninvasive EEG~\citep{Guger2017}.  However, methods of decoding signals based on the detection of waveforms has been scarce. Hence, this traditional and knowledgeable approach has been neglected particularly in BCI Research. We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and which can decode them in a supervised and semi-automated procedure.

@@

@@

Farwell and Donchin P300 Speller \citep{Farwell1988} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.  This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.  This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.

@@

The template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.  This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.

An EEG stream with null-P300 signal is obtained by the following procedure: 
A subject participant is recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and she/he agrees and signs a written informed consent.  This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.  The participant is healthy and have normal or corrected-to-normal vision and no history of neurological disorders. This voluntary subject is aged between 20-30 years old.  EEG data is collected in a single recording session. She/He is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.  The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.  Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is set to 250 Hz.

The participant is instructed to passively watch the flashing screen while not focusing on any particular letter.  A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.

@@

@@

Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.  Table \ref{tab:results} shows the  performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance is attained.   Figure \ref{fig:performancetest} shows the performance curves for six algorithms.  Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.  As previously described the data is continuously divided in two sets, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance. This is repeated one hundred times, and performances averaged.   Figure \ref{fig:performancetestlatency} shows the same results for the Experiment 2, where a noisy latency lag was included.   Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.  Furthermore, results obtained for the dataset BCI Competition 2003 IIb are shown in Figures \ref{fig:performancebcicompetition} and in Table~\ref{tab:bcicompetitionresults}.  For this experiment the number of available intensification sequences is 15.

@@

@@

@@

Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.  Results for this competition show perfect classification with tailored algorithm.  Even comparing the results with SVM multichannel, which uses all the available information from all channels show comparable results.

It is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.  It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.

@@

The conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.   At the same time, it is a subjective time-consuming task, with long-learning curves, requires specialized personnel, and it has significant error rates~\citep{Tjepkema-Cloostermans2018}.  These problems has pushed for the adoption of more automated means of decoding the signals~\citep{Thakor2004}.   This trend pointed to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}~\citep{Wulsin2011,Tjepkema-Cloostermans2018}.

@@

These methods have the advantage that they can map a visual component with a clinical meaning to a feature with an objective representation. Compound classifiers or ensemble of features can be explored.  Successful approaches in Computer Vision or Pattern Recognition in other areas use them~\citep{Criminisi2013} with a significant enhancement of the classification performances~\citep{Gu2012}.

>>>>

The need of quantitative procedures to automate the decoding of EEG signals has been materialized in BCI where around $71.2\%$ is based on noninvasive EEG~\citep{Guger2017}.  However, methods of decoding signals based on the detection of waveforms has been scarce. Hence, the traditional and knowledgeable approach has been neglected particularly in BCI Research. We aim to help fix this gap by providing a review of the methods which emphasize the waveform, the shape of the EEG signal and which can decode them in a supervised and semi-automated procedure.

@@

Table~\ref{tab:methods} shows a list of depictions used to describe waveforms in the surveyed literature.  Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.  For ERPs on the other hand, more elaborate indexes has been provided establishing relations between signal components amplitudes.  Finally, in sleep studies and ICU research areas is where the most complex indexes has been derived, particularly coupling signal properties like phase, amplitude and frequency.

@@

Farwell and Donchin P300 Speller \citep{Farwell1988,Rakotomamonjy2008} is one the most used BCI paradigms to implement a thought translation device and to send commands to a computer in the form of selected letters, similar to typing on a virtual keyboard.  This procedure exploits a cognitive phenomena raised by the oddball paradigm: along the EEG trace of a person which is focusing on a sequence of two different visual flashing stimulus, a particular and distinctive transient component is found each time the expected stimulus flashes.  This is cleverly utilized in the P300 Speller, where rows and columns of a 6x6 matrix flashes randomly but only the flashing of a column or row where the letter that a user is focusing will trigger concurrently the P300 ERP along the EEG trace.

@@

\paragraph{Passive Modality}

In this case the template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.  This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.  This template is selected due to its shape which more closely resemble the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.

An EEG stream with null-P300 signal is obtained by the following procedure: 
Four (4) participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the Declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and they agree and sign a written informed consent.  This study is approved by the \textit{Departamento de Investigación y Doctorado, Instituto Tecnológico de Buenos Aires (ITBA)}.  The participants are healthy and have normal or corrected-to-normal vision and no history of neurological disorders. These voluntary subjects are aged between 20-40 years old.  EEG data is collected in a single recording session. Each subject is seated in a comfortable chair, with her/his vision aligned to a computer screen located one meter in front of her/him.  The handling and processing of the data and stimuli is conducted by the OpenVibe platform~\citep{Renard2010}.  Gel-based active electrodes (g.LADYbird, g.Tec, Austria) are used on locations Fz, Cz, Pz, Oz, P3,P4, PO7 and PO8 according to the 10-20 international system.  Reference is set to the right ear lobe and ground is preset as the AFz position.   Sampling frequency is set to 250 Hz.

All participants are instructed to passively watch the flashing screen while not focusing on any particular letter.  None of them have any experience with a BCI device.  A questionnaire is handed out at the end of the experiment with questions about how the participant felt during it, without giving more details.

@@

\paragraph{Active modality}

The experiment conducted by~\citep{Riccio2013} is replicated on 4 healthy subjects, performing a copy-spelling task with feedback.  In this case, to construct the pseudo-real dataset each template must first be obtained from each subject.  The most prototypical P300 waveform template is selected by visual observation as well.  This template is superimposed to each segment marked as hit on the EEG stream.  The baseline of each segment, on the other hand, is obtained by picking randomly any no hit segment for the same subject.

@@

Results are shown in Table \ref{tab:results} and in Figure \ref{fig:performancetest},\ref{fig:performancetestlatency} and \ref{fig:performancetestamplitude}.  Table \ref{tab:results} shows the  performance while identifying each letter of the standard P300 Speller Matrix, and the channel where the best performance is attained.   Figure \ref{fig:performancetest} shows the performance curves for six algorithms.  Each one represents the percentage of letters that is actually predicted by the algorithms using a cross-validation procedure.  As previously described the data is continuously divided in two sets, where the first 15 letters are used to derive the dictionary of templates while the remaining 20 letters are used to measure the letter identification performance. This is repeated one hundred times, and performances averaged.

@@

Figure \ref{fig:performancetestlatency} shows the same results for the Experiment 2, where a noisy latency lag was included.   Finally, Figure \ref{fig:performancetestamplitude} represents the performance values obtained for the Experiment 3, when the amplitude of the P3b component of the template is randomly attenuated.

@@

Furthermore, results obtained for the dataset BCI Competition 2003 IIb are shown in Figures \ref{fig:performancebcicompetition} and in Table~\ref{tab:bcicompetitionresults}.  For this experiment the number of available intensification sequences is 15.

@@

Regarding results produced for the public and real dataset IIb of P300 ERP from the Berlin BCI Competition II (2003), the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of $70\%$~\citep{Kathner2017}.  When the character identification rate reaches this level of performance, the usage of word predicting algorithms allows to implement practical speller applications.  Results for this competition have shown perfect classification with tailored algorithm.  This level is also similar to the performance obtained for the Experiment 3, which represents coincidentally the more realistic scenario for the pseudo-real dataset.  It is important to remark that the algorithms presented here analyze the waveform structure of a single-channel signal~\citep{Rakotomamonjy2008,Gribonval2008}.

@@

%The conventional clinical method of observing the waveform is understood to be subjective and laborious because results depend on the technicians' experience and expertise.   At the same time, it is a subjective time-consuming task, with long-learning curves, requires specialized personnel, and it has significant error rates~\citep{Tjepkema-Cloostermans2018}.  These problems has pushed for the adoption of more automated means of decoding the signals~\citep{Thakor2004}.   This trend pointed to the initial development of quantitative EEG, which however didn't replaced clinically the traditional approach which is still widespread: the Gold standard in clinical EEG is still \textit{Eye Ball}~\citep{Wulsin2011,Tjepkema-Cloostermans2018}.

@@

%These methods have the advantage that they can map a visual component with a clinical meaning to a feature with an objective representation. Compound classifiers or ensemble of features can be explored.  Successful approaches in Computer Vision or Pattern Recognition in other areas use them~\citep{Criminisi2013} with a significant enhancement of the classification performances~\citep{Gu2012}.
